Quantum Hoare Type Theory

### 2020-10-30 ms-paper/ms-paper.tex
Use classic(X) everywhere not classical(X)
A typo and unneeded statement

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{classical}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

@@

\paragraph{Propositions}
The predicate language is inspired by \citet{unruh2019} that follows subspace semantics. We need to be explicit about what we mean when we say that a certain state satisifes a certain predicate in the quantum setting. In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state. We use the metavariable \textbf{X} to refer to a list of quantum variables. Here are some examples:
\begin{itemize}
	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
	\item $\bot$ represents nothing, it is never satisfied.
	\item $P \wedge Q$ represent that both P and Q hold (conjunction), it is equivalent to the intersection of the subspaces.
	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
	\item If a unitary operator U is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
	\item If we want to write that a list of variables \textbf{X} is in a specific state, we can model that proposition as the variables being in the span of the state and use the notation, $\textbf{X} =_q \psi$.
\end{itemize}

@@

\paragraph{Verification}
We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function. For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below. We have elided specific details about unitary application as they are still being worked out.
\newpage
\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
              -- \top
         = do qa <= init false;
              -- qa =q |0\>, class(qa)
              apply H to qa;
              -- qa =q |+\>, uniform(qa)
              qb <= init false;
              -- qa =q |+\>, uniform(qa), qb =q |0\>, class(qb)
              apply CNOT to (qa, qb);
              -- (qa, qb) =q |++\>  -- can use a weaker but more general `EPR' here
              -- /\ uniform(qa, qb) -- this won't apply then
              a = meas qa;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true
              b = meas qb;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false, b = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true,  b = true
              return (a, b)
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\todo{TODO: how is classical equality ensured?}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
         = do (qa, qb) <- bell00;
              a <= meas qa;
              b <= meas qb;
              return (a, b)
\end{lstlisting}

>>>>

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Propositions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid U \cdot P \mid \textbf{X} =_q \psi \mid \kw{uniform}(\textbf{X}) \mid \kw{separable}(\textbf{X}) \mid \kw{class}(\textbf{X}) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.5\columnwidth}%
			$ \kw{\textbf{init}} N \mid \kw{\textbf{apply}} U \kw{\textbf{to}} N \mid \kw{\textbf{meas}} N$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x = N; E $\\ \\
%		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
%		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
%		\textit{Proposition context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
		\textit{Unitaries} & $U$ & ::= & $ I \mid X \mid Y \mid Z \mid H \mid CX \mid CZ \mid S \mid T$
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

@@

\paragraph{Propositions}
The predicate language is inspired by \citet{unruh2019} that follows subspace semantics. We need to be explicit about what we mean when we say that a certain state satisfies a certain predicate in the quantum setting. In this model, the propositions are subspaces of the Hilbert space corresponding to a quantum state. We use the metavariable \textbf{X} to refer to a list of quantum variables. Here are some examples:
\begin{itemize}
	\item $\top$ represents the complete Hilbert space that is always satisfied by any state.
	\item $\bot$ represents nothing, it is never satisfied.
	\item $P \wedge Q$ represent that both P and Q hold (conjunction), it is equivalent to the intersection of the subspaces.
	\item $P \vee Q$ similarly represents disjunction and is modeled as the sum (linear combination) of the subspaces P and Q.
	\item If a unitary operator U is applied to a state satisfying a proposition P, then the new proposition can be written as $U \cdot P$.
	\item If we want to write that a list of variables \textbf{X} is in a specific state, we can model that proposition as the variables being in the span of the state and use the notation, $\textbf{X} =_q \psi$.
\end{itemize}

@@

\paragraph{Verification}
We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function. For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below.
\newpage
\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
              -- \top
         = do qa <= init false;
              -- qa =q |0\>, class(qa)
              apply H to qa;
              -- qa =q |+\>, uniform(qa)
              qb <= init false;
              -- qa =q |+\>, uniform(qa), qb =q |0\>, class(qb)
              apply CNOT to (qa, qb);
              -- (qa, qb) =q |++\>  -- can use a weaker but more general `EPR' here
              -- /\ uniform(qa, qb) -- this won't apply then
              a = meas qa;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true
              b = meas qb;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false, b = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true,  b = true
              return (a, b)
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
         = do (qa, qb) <- bell00;
              a <= meas qa;
              b <= meas qb;
              return (a, b)
\end{lstlisting}
